State and Parameter Estimation for Spatio-Temporal Neural Fields

### 2010-06-11 ltx/Journal of Neuroengineering/StateAndParameterEstimationNeuralFields_JoNE.tex
corrected some little typos. Re did the symbol table as a bunch of little symbol tables as I think it looks a bit nicer...

% \begin{document}

@@

\begin{figure}[ th]\label{fig:2d_kernel}

@@

   	\caption{A mexican-hat connectivity kernel, $w(\mathbf{r},\mathbf{r'})$ used in this paper. The function $w(.)$ is rotationally symetric (isotropic) about zero, hence a cross section captures the important aspect of the connectivity kernel's shape. The kernel decays asymptotically to zero.} 

@@

\section{Results}\label{ResultsSection} The non-reduced neural field model derived in Section~\ref{NeuralModelSection} was used to generate the data for state and parameter estimation. All parameters for the model are given in Table~\ref{tab:Model Parameters}. Figure~\ref{fig:experimental design}(a) shows example data from the sensors of the model, and figure~\ref{fig:experimental design}(b) shows the power spectral density (PSD) of the simulated time-series. The PSD shows the typical $1/f^2$ characteristics of intracranial EEG.

@@

\caption{Parameters for neural field model used to generate data to demonstrate the estimation framework.}

@@

   	\caption{Example of experimental design and model output.  \textbf{a} Example of data generated by the non-reduced model through a subset of the observations. \textbf{b}. Power spectral density of the data generated by an observation showing the typical $1/f^2$ characteristic of iEEG.} 

@@

@@

\caption{Spatial frequency of the neural fields. \textbf{a.} The average (over time) spatial frequency of the neural field. \textbf{b.} The average (over time) spatial frequency of the estimated neural field resulting from the basis function decomposition. As a result of the band limiting properties of the sensors and basis functions the reconstructed field shows a narrower bandwidth.}

@@

\caption{Distribution of the parameters estimates over 100

@@

The frequency analysis and model selection tools presented in this paper specify the sensor and field basis function arrangement required to form a state-space model that is capable of capturing the dominant cortical dynamics. In addition to providing conditions to ensure the reduced model can capture the significant spectral properties of the field, these new tools can be used to better design electrophysiological recording techniques avoid spatial aliasing. This analysis is becoming increasingly relevant as recording systems are becoming more sophisticated with higher density recordings~\cite{Brinkmann2009}. The effect of the basis function decomposition will restrict the frequency content of the estimated field. This is evident in when comparing Figures~\ref{fig:FieldFFT} and~\ref{fig:EstimatedFieldFFT} where the estimated field is spatially band-limited. The spatial frequency content of the field is governed by disturbance covariance and the shape of the connectivity kernel.
% 
% In practice Even though the some high spatial frequency content of the connectivity kernel is lost through the basis function decomposition, the dominant dynamics are still captured with the estimated reduced model. This is illustrated in figure~\ref{fig:FieldEstimate}. In addition, the ratio of the kernel parameters follows a tight distribution as seen in figure~\ref{fig:ParametersRatio}. This indicates that the ratio of excitation to inhibition can be tracked with this approach.

@@

\section{Parameters} 
\begin{tabular}
	{c|c} Symbol & Description \\
	\hline
	\\ & \textbf{Domain and indicies} \\
	\hline
	$\Omega$ & spatial domain \\
	$t$ & time (seconds) \\
	$\mathbf{r}$ & spatial location \\
	$n$ & sensor index $n=1,..,N$ \\
	$T$ & time step \\
	\\ & \textbf{Spatiotemporal Signals} \\
	\hline
	$y(\mathbf{r}_n,t)$ & observation \\
	$v(\mathbf{r},t)$ & mean membrane potential \\
	$g(\mathbf{r},t)$ & average action potential rate \\
	$f(\mathbf{r},t)$ & firing function rate \\
	$u(\mathbf{r},t)$ & external input \\
	$e(\mathbf{r},t)$ & field disturbance, covariance $\gamma$\\
	$\epsilon(\mathbf{r}_n,t)$ & observation noise, covariance $\Sigma_\epsilon$ \\
	\\ & \textbf{Model} \\
	\hline
	$h(t)$ & post-synaptic response kernel \\
	$m(\mathbf{r},\mathbf{r}')$ & sensor kernel, variance $\sigma_m^2$ \\
	$\eta(t)$ & Heaviside function \\
	$\zeta$ & inverse synaptic time constant \\
	$w(\mathbf{r},\mathbf{r}')$ & spatial connectivity kernel \\
	$f_{max}$ & maximal firing rate \\
	$\varsigma$ & slope of sigmoidal activation function \\
	$v_0$ & firing threshold \\
	$\delta(t)$ & Dirac-delta function \\
	$D$ & Temporal differential operator \\
	$\xi$ & time constant parameter \\
	\\ & \textbf{Decomposition} \\
	\hline
	$L$ & number of field basis functions  \\
	$\mathbf{\phi(r)}$ & vector of Gaussian basis functions \\
	$\mathbf{x}_t$ & state vector at time $t$ \\
	$\mathbf{\psi}$ & vector of connectivity kernel basis functions \\
	$\theta$ & vector of connectivity kernel parameters \\
	$\Gamma$ & inner product of field basis functions \\
	$q()$ & state function \\
	$k()$ & maps into to discrete state of input function \\
	$\mathbf{e}_t$ & state disturbance, covariance $\Sigma_e$ \\
	$\mathbf{C}$ & observation matrix \\
	\\ & \textbf{Frequency Analysis} \\
	\hline
	$V(\nu)$ & Spectrum of the dynamic field \\
	$\nu$ & Spatial frequency \\
	$\nu_c$ & Spatial cut-off frequency \\
	$\Delta_s$ & Distance between adjacent sensors \\
	$\rho$ & Over-sampling parameter \\
	$\Delta_b$ & Distance between field basis functions \\
	$\sigma_{nu}^2$ & Variance of FT of Gaussian basis function \\
	$\sigma^2$ & Spatial variance of Gaussian in spatial domain \\
	\\ & \textbf{State Estimation} \\
	\hline
	$\hat{\mathbf{x}}$ & State estimate \\
	$\chi$ & Matrix of sigma vectors \\
	$\hat{\mathbf{x}}_t^{f-}$ & Forward prior estimate \\
	$\hat{\mathbf{x}}_t^f$ & Forward posterior estimate \\
	$\hat{\mathbf{x}}_t^{b-}$ & Backward prior estimate \\
	$\hat{\mathbf{x}}_t^{b}$ & Backward posterior estimate \\
	$P^f_t$ & Forward posterior covariance matrix \\
	$P^{f-}_t$ & Forward prior covariance matrix \\
\end{tabular}
\section*{References} 
\bibliographystyle{unsrt} 
\bibliography{BrainIDE}

>>>>

@@

\begin{figure}[th]\label{fig:2d_kernel}

@@

   	\caption{A mexican-hat connectivity kernel, $w(\mathbf{r},\mathbf{r'})$ used in this paper. The function $w(.)$ is rotationally symmetric (isotropic) about zero, hence a cross section captures the important aspect of the connectivity kernel's shape. The kernel decays asymptotically to zero with space.} 

@@

\section{Results}\label{ResultsSection} The non-reduced neural field model derived in Section~\ref{NeuralModelSection} was used to generate the data for state and parameter estimation. All parameters for the model are given in Table~\ref{tab:Model Parameters}. Figure~\ref{fig:experimental design}(a) shows example data from the sensors of the model, and Figure~\ref{fig:experimental design}(b) shows the power spectral density (PSD) of the simulated time-series. The PSD shows the typical $1/f^2$ characteristics of intracranial EEG.

@@

\caption{Parameters for the neural field model, which are used to generate synthetic data.}

@@

   	\caption{\textbf{a.} Example of data generated by the non-reduced model. \textbf{b.} Power spectral density of the data generated by an observation showing the typical $1/f^2$ characteristic of iEEG.} 

@@

\begin{figure}

@@

\caption{Spatial frequency of the neural fields. \textbf{a.} The average (over time) spatial frequency of the neural field. \textbf{b.} The average (over time) spatial frequency of the estimated neural field. As a result of the band limiting properties of the sensors and basis functions the reconstructed field shows a narrower bandwidth.}

@@

\caption{Distribution of the parameter estimates over 100

@@

The frequency analysis and model selection tools presented in this paper specify the sensor and field basis function arrangement required to form a state-space model that is capable of capturing the dominant cortical dynamics. In addition to providing conditions to ensure the reduced model can capture the significant spectral properties of the field, these new tools can be used to better design electrophysiological recording techniques avoid spatial aliasing. This analysis is becoming increasingly relevant as recording systems are becoming more sophisticated with higher density recordings~\cite{Brinkmann2009}. The effect of the basis function decomposition will restrict the frequency content of the estimated field. This is evident in when comparing Figures~\ref{fig:FieldFFT} and~\ref{fig:EstimatedFieldFFT} where the estimated field is spatially band-limited. The spatial frequency content of the field is governed by disturbance covariance and the shape of the connectivity kernel.
% 
% In practice Even though the some high spatial frequency content of the connectivity kernel is lost through the basis function decomposition, the dominant dynamics are still captured with the estimated reduced model. This is illustrated in Figure~\ref{fig:FieldEstimate}. In addition, the ratio of the kernel parameters follows a tight distribution as seen in Figure~\ref{fig:ParametersRatio}. This indicates that the ratio of excitation to inhibition can be tracked with this approach.

@@

\section{Notation}

\begin{table}[h!]\footnotesize
    \centering
    \begin{tabular}{cl}
        $\Omega$ & spatial domain \\
	    $t$ & time (seconds) \\
    	$\mathbf{r}$ & spatial location \\
	    $n$ & sensor index $n=1,..,N$ \\
    	$T$ & time step
    \end{tabular}
    \caption{Domain and indices}
\end{table}

\begin{table}[h!]\footnotesize
    \centering
    \begin{tabular}{cl}
        $y(\mathbf{r}_n,t)$ & observation \\
	    $v(\mathbf{r},t)$ & mean membrane potential \\
    	$g(\mathbf{r},t)$ & average action potential rate \\
    	$f(\mathbf{r},t)$ & firing function rate \\
    	$e(\mathbf{r},t)$ & field disturbance, covariance $\gamma$\\
    	$\epsilon(\mathbf{r}_n,t)$ & observation noise, covariance$\Sigma_\epsilon$ \\
    \end{tabular}
    \caption{Spatiotemporal Signals}
\end{table}

\begin{table}[h!]\footnotesize
    \centering
    \begin{tabular}{cl}
        	$h(t)$ & post-synaptic response kernel \\
	$m(\mathbf{r},\mathbf{r}')$ & sensor kernel, width $\sigma_m^2$ \\
	$\eta(t)$ & Heaviside function \\
	$\zeta$ & inverse synaptic time constant \\
	$w(\mathbf{r},\mathbf{r}')$ & spatial connectivity kernel \\
	$f_{max}$ & maximal firing rate \\
	$\varsigma$ & slope of sigmoidal activation function \\
	$v_0$ & firing threshold \\
	$\delta(t)$ & Dirac-delta function \\
	$D$ & Temporal differential operator \\
	$\xi$ & time constant parameter
    \end{tabular}
    \caption{Model}
\end{table}

\begin{table}[h!]\footnotesize
    \centering
    \begin{tabular}{cl}
        $L$ & number of field basis functions  \\
    	$\mathbf{\phi(r)}$ & vector of Gaussian basis functions \\
    	$\mathbf{x}_t$ & state vector at time $t$ \\
    	$\mathbf{\psi}$ & vector of connectivity kernel basis functions \\
    	$\theta$ & vector of connectivity kernel parameters \\
    	$\Gamma$ & inner product of field basis functions \\
    	$q()$ & state function \\
    	$k()$ & maps into to discrete state of input function \\
    	$\mathbf{e}_t$ & state disturbance, covariance $\Sigma_e$ \\
    	$\mathbf{C}$ & observation matrix \\
    \end{tabular}
    \caption{Decomposition}
\end{table}

\begin{table}[h!]\footnotesize
    \centering
    \begin{tabular}{cl}
        $V(\nu)$ & Spectrum of the dynamic field \\
	$\nu$ & Spatial frequency \\
	$\nu_c$ & Spatial cut-off frequency \\
	$\Delta_s$ & Distance between adjacent sensors \\
	$\rho$ & Over-sampling parameter \\
	$\Delta_b$ & Distance between field basis functions \\
	$\sigma_{nu}^2$ & Variance of FT of Gaussian basis function \\
	$\sigma^2$ & Spatial variance of Gaussian in spatial domain \\
    \end{tabular}
    \caption{Frequency Analysis}
\end{table}

\begin{table}[h!]\footnotesize
    \centering
    \begin{tabular}{cl}
        $\hat{\mathbf{x}}$ & State estimate \\
	    $\chi$ & Matrix of sigma vectors \\
    	$\hat{\mathbf{x}}_t^{f-}$ & Forward prior estimate \\
    	$\hat{\mathbf{x}}_t^f$ & Forward posterior estimate \\
    	$\hat{\mathbf{x}}_t^{b-}$ & Backward prior estimate \\
    	$\hat{\mathbf{x}}_t^{b}$ & Backward posterior estimate \\
    	$P^f_t$ & Forward posterior covariance matrix \\
    	$P^{f-}_t$ & Forward prior covariance matrix
    \end{tabular}
    \caption{State Estimation}
\end{table}

\newpage

\section*{References} 
\bibliographystyle{unsrt} 
\bibliography{BrainIDE}
